---
title: "Protein DB"
author: "Ghanyy, 106080"
date: "19 listopada 2015"
output: html_document
---


# Introduction

Purpose of this project is to analyse data from [Protein Data Bank](http://www.rcsb.org/) (PDB) using some basic [R](https://www.r-project.org/) functionalities.

# Table of contents

1. [Libraries](#libraries)
2. [Initialisation](#initialisation)
3. [Data preparation](#data-preparation)
4. [Data explanation](#data-explanation)
5. [Variables correlation](#variables-correlation)
6. [Some statistics](#some-statistics)
  * [Number of examples per `res_name`](#number-of-examples-per-res_name)
  * [Distribution of atom numbers and electron numbers](#distribution-of-atom-numbers-and-electron-numbers) 
  * [Heatmap based on atoms and electrons distributions](#heatmap-based-on-atoms-and-electrons-distributions)
  * [Ten classes with most inconsonant atom number](#ten-classes-with-most-inconsonant-atom-number)
  * [Ten classes with most inconsonant electron number](#ten-classes-with-most-inconsonant-electron-number)
  * [Distribution of all columns with `part_01` prefix](#distribution-of-all-columns-with-part_01-prefix)
  
## Libraries

Following libraries were used:
```{r libraries, message=FALSE}
library(knitr) # provides a general-purpose tool for dynamic report generation.
library(dplyr) # data manipulation tools for working with data frames.
library(ggplot2) # powerful graphics language for creating elegant and complex plots.
library(gtable) # marginal histograms in ggplot2.
library(reshape2) # allows flexibly restructure and aggregate data using just two functions: melt and cast.
library(corrplot) # used for displaying correlation matrix.
```

## Initialisation

  In order to obtain the same results every time seed is fixed:
```{r initialisation, cache=TRUE}
set.seed(106080)
```

## Data preparation

  1. Load raw data from file:
```{r data-preparation, cache=TRUE}
fileDir <- paste(getwd(), "all_summary.txt", sep = "/") # data should be in the same directory as this script
PdbRawData <- read.csv(file = fileDir, dec = ".", sep = ";", header = TRUE, na.strings = c("nan"), encoding = "UTF-8", fileEncoding = "UTF-8")
```
  2. Delete rows unnecessary for this project:
    * `res_name` != `DA`, `DC`, `DT`, `DU`, `DG`, `DI`, `UNK`, `UNX`, `UNL`, `PR`, `PD`, `Y1`, `EU`, `N`, `15P`, `UQ`, `PX4`, `NAN`
    * Leave only unique pairs of values `pdb_code`, `res_name`

```{r data-preparation2, cache=TRUE}
filterValuesVector <- c("DA", "DC", "DT", "DU", "DG", "DI", "UNK", "UNX", "UNL", "PR", "PD", "Y1", "EU", "N", "15P", "UQ", "PX4")
PdbFilteredData <- PdbRawData %>% 
filter(!(res_name %in% filterValuesVector) | is.na(res_name)) %>% 
distinct(pdb_code, res_name)
```
  3. Delete columns with only `NA` values:
```{r data-preparation3, cache=TRUE}
filterNAVector <- !sapply(PdbFilteredData, function(x)all(is.na(x)))
PdbFilteredData <- PdbFilteredData[filterNAVector]
```
## Data explanation

After initial filtering our dataset have following properties:

[Next section...](#variables-correlation)

  1. Column classes and sample:
```{r data-explanation, cache=TRUE}
str(PdbFilteredData)
```
  2. Summary:
```{r data-explanation2, cache=TRUE}
summary.data.frame(PdbFilteredData)
```

## Variables correlation

All plots uses significance test:
```{r variables-correlation-sig-test, cache=TRUE}
cor.mtest <- function(mat, conf.level = 0.95) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat <- lowCI.mat <- uppCI.mat <- matrix(NA, n, n)
    diag(p.mat) <- 0
    diag(lowCI.mat) <- diag(uppCI.mat) <- 1
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], conf.level = conf.level)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
            lowCI.mat[i, j] <- lowCI.mat[j, i] <- tmp$conf.int[1]
            uppCI.mat[i, j] <- uppCI.mat[j, i] <- tmp$conf.int[2]
        }
    }
    return(list(p.mat, lowCI.mat, uppCI.mat))
}
```

  1. Correlation between columns with prefix `local`, `local_min` was excluded due to `NA` values:
  
```{r variables-correlation, out.width = 900, cache=TRUE}
localColumns <- PdbFilteredData %>% select(starts_with("local"), -local_min)
localColumns <- localColumns[sapply(localColumns, is.numeric)]
correlationMatrix <- cor(localColumns, use = "complete.obs", method = "pearson")
sigTest <- cor.mtest(localColumns, 0.95)

corrplot(correlationMatrix, order = "AOE", tl.cex = 0.5, tl.col = "black", p.mat = sigTest[[1]], sig.level = 0.05, insig = "blank", cl.pos = "b")
```

  2. Correlation between remaining columns, prefix `part` and `dict` were excluded due to high correlation, `solvent_radius`, `solvent_opening_radius`, `resolution_max_limit` and `grid_space` were excluded due to `NA`:

```{r variables-correlation2, out.width = 1200}
remainingColumns <- PdbFilteredData %>% select(-starts_with("part"), -starts_with("dict"), -starts_with("local"), -solvent_radius, -solvent_opening_radius, -resolution_max_limit, -grid_space)
remainingColumns <- remainingColumns[sapply(remainingColumns, is.numeric)]
correlationMatrix <- cor(remainingColumns, use = "complete.obs", method = "pearson")
sigTest <- cor.mtest(remainingColumns, 0.95)
corrplot(correlationMatrix, order = "AOE", tl.cex = 0.4, tl.col = "black", p.mat = sigTest[[1]], sig.level = 0.05, insig = "blank", cl.pos = "b")
```

## Some statistics
### Number of examples per `res_name`


```{r stats, cache=TRUE}
resNameCount <- PdbFilteredData %>% select(res_name) %>% group_by(res_name) %>% summarise(count = n()) %>% arrange(desc(count), res_name)
head(resNameCount, 20)
```
```{r stats2, out.width = 900, echo=FALSE, cache=TRUE}
resNamePlot <- resNameCount %>% select(count) %>% arrange(count)
ggplot(resNameCount, aes(x = count)) + 
xlab("count(res_name)") +
ylab("density") +
geom_density(kernel = "gaussian", fill = "steelblue") + 
scale_y_sqrt() + 
theme_minimal()
```

### Distribution of atom numbers and electron numbers

  
```{r stats3, out.width = 900, cache=TRUE, message=FALSE}
atomsAndElectrons <- PdbFilteredData %>% 
select(local_res_atom_non_h_count, local_res_atom_non_h_electron_sum) %>% 
transmute(atoms = local_res_atom_non_h_count, electrons = local_res_atom_non_h_electron_sum)
atomsAndElectrons <- melt(atomsAndElectrons)

ggplot(atomsAndElectrons) + 
geom_freqpoly(aes(x = value, y = ..density.., colour = variable), binwidth = 1) + 
xlab("count(variable)") + 
scale_x_sqrt() +
theme_minimal()
```

### Heatmap based on atoms and electrons distributions

  * X - `local_res_atom_non_h_count`, Y - `local_res_atom_non_h_electron_sum`

```{r stats4, out.width = 650, message=FALSE, cache=TRUE}
atomsAndElectrons2 <- PdbFilteredData %>% 
select(local_res_atom_non_h_count, local_res_atom_non_h_electron_sum) %>% 
transmute(atoms = local_res_atom_non_h_count, electrons = local_res_atom_non_h_electron_sum)
atomsAndElectrons2 <- melt(atomsAndElectrons2, id.vars=c("atoms", "electrons"))

secretPlot <- ggplot(atomsAndElectrons2, aes(x = electrons, y = atoms)) +
stat_density2d(geom="tile", aes(fill = ..density..), contour = FALSE, binwidth = 1) +
coord_cartesian(xlim = c(0,650), ylim = c(0,100)) +
scale_x_continuous(breaks=seq(100, 700, 100)) +
scale_y_continuous(breaks=seq(20, 100, 20)) +
scale_fill_distiller(palette = "Spectral", trans = "reverse")  +
theme_classic() +
theme(legend.position="none", axis.title=element_blank(), axis.line.y=element_blank(), text=element_text(size=16, family="Liberation Sans"), axis.text.y=element_text(angle=90, hjust=0.5))

# tutorial - http://sandymuspratt.blogspot.co.uk/2013/02/scatterplot-with-marginal-boxplots.html
theme_remove_all <- theme(axis.text = element_blank(),
  line = element_blank(),
  panel.background = element_blank(),
  panel.margin = unit(0, "lines"),
  axis.title = element_blank(),
  axis.ticks =  element_blank(),
  axis.ticks.margin = unit(0, "lines"),
  axis.ticks.length = unit(0, "cm"))

xMargin <- ggplot(atomsAndElectrons2, aes(x = electrons)) +
geom_histogram(binwidth=6.5, fill="red", color="black") +
scale_x_continuous(breaks=seq(0, 700, 100)) +
theme_remove_all +
theme(plot.margin= unit(c(-0.5, 0, 0, -0.5), "lines"))

yMargin <- ggplot(atomsAndElectrons2, aes(x = atoms)) +
geom_histogram(binwidth=1, fill="red", color="black") +
scale_x_continuous(breaks=seq(0, 100, 20)) +
coord_flip() +
theme_remove_all +
theme(plot.margin= unit(c(0, -0.5, -0.5, 0), "lines"))

# Get the gtables
gt1 <- ggplot_gtable(ggplot_build(secretPlot))
gt2 <- ggplot_gtable(ggplot_build(xMargin))
gt3 <- ggplot_gtable(ggplot_build(yMargin))

# Get maximum widths and heights for x-axis and y-axis title and text
maxWidth = unit.pmax(gt1$widths[2:3], gt2$widths[2:3])
maxHeight = unit.pmax(gt1$heights[4:5], gt3$heights[4:5])

# Set the maximums in the gtables for gt1, gt2 and gt3
gt1$widths[2:3] <- as.list(maxWidth)
gt2$widths[2:3] <- as.list(maxWidth)

gt1$heights[4:5] <- as.list(maxHeight)
gt3$heights[4:5] <- as.list(maxHeight)

# Combine the scatterplot with the two marginal boxplots
# Create a new gtable
gt <- gtable(widths = unit(c(7, 2), "null"), height = unit(c(2, 7), "null"))

# Instert gt1, gt2 and gt3 into the new gtable
gt <- gtable_add_grob(gt, gt1, 2, 1)
gt <- gtable_add_grob(gt, gt2, 1, 1)
gt <- gtable_add_grob(gt, gt3, 2, 2)

# And render the plot
grid.newpage()
grid.draw(gt)
```

### Ten classes with most inconsonant atom number


```{r stats5, cache=TRUE}
top10 <- PdbFilteredData %>% 
select(class=res_name, local_res_atom_non_h_count, dict_atom_non_h_count) %>%
mutate(difference=abs(local_res_atom_non_h_count - dict_atom_non_h_count)) %>%
arrange(desc(difference)) %>%
distinct(class) %>%
slice(1:10)

atomInconsonance <- PdbFilteredData %>% 
select(class=res_name, local_res_atom_non_h_count, dict_atom_non_h_count) %>%
mutate(difference=abs(local_res_atom_non_h_count - dict_atom_non_h_count)) %>%
filter(class %in% top10$class) %>%
group_by(class) %>%
summarise(minimum = min(difference),
          maximum = max(difference),
          mean = mean(difference),
          median = median(difference),
          variation = var(difference),
          std_dev = sd(difference)) %>%
ungroup() %>%
arrange(desc(maximum))

kable(atomInconsonance)

```

### Ten classes with most inconsonant electron number


```{r stats6, cache=TRUE}
top10 <- PdbFilteredData %>% 
select(class=res_name, local_res_atom_non_h_electron_sum, dict_atom_non_h_electron_sum) %>%
mutate(difference=abs(local_res_atom_non_h_electron_sum - dict_atom_non_h_electron_sum)) %>%
arrange(desc(difference)) %>%
distinct(class) %>%
slice(1:10)

electronInconsonance <- PdbFilteredData %>% 
select(class=res_name, local_res_atom_non_h_electron_sum, dict_atom_non_h_electron_sum) %>%
mutate(difference=abs(local_res_atom_non_h_electron_sum - dict_atom_non_h_electron_sum)) %>%
filter(class %in% top10$class) %>%
group_by(class) %>%
summarise(minimum = min(difference),
          maximum = max(difference),
          mean = mean(difference),
          median = median(difference),
          variation = var(difference),
          std_dev = sd(difference)) %>%
ungroup() %>%
arrange(desc(maximum))

kable(electronInconsonance)

```

### Distribution of all columns with `part_01` prefix

```{r stats7, cache= TRUE, warning=FALSE}
partColumns <- PdbFilteredData %>% 
select(starts_with("part_01"))

for(colName in names(partColumns)) {
  currCol <- partColumns[[colName]]
  currColNoNA <- currCol[!is.na(currCol)]
  currMean <- mean(currColNoNA)
  currPlot <- ggplot(partColumns, aes(x = currCol)) + 
  xlab(paste("count(", colName, ")")) +
  ylab("density") +
  geom_density(kernel = "gaussian", fill = "steelblue") + 
  theme_minimal() +
  geom_rug(sides="b", color="blue", size = 2, aes(x=currMean, y=0)) + 
  geom_text(aes(label=round(currMean, 2), x=currMean, y=0), hjust=0.5, vjust=-1, color="blue")
  print(currPlot)
}
```
